Physics-based protein structure prediction and design using the confinement method

### 2013-01-11 Paper/paper.tex
changed paper 11_6

% Custom macros for author comments
\newcommand{\Alberto}[1]{\color{ForestGreen}#1\normalcolor }
\newcommand{\Justin}[1]{\color{blue}#1\normalcolor}
\newcommand{\Arijit}[1]{\color{yellow}#1\normalcolor}
\newcommand{\Ken}[1]{\color{red}#1\normalcolor}

@@

In computational structural biology there are numerous cases where the free energy difference
between two well defined 
states plays an important role. Examples range from small to large conformational changes of proteins due to
ligand binding, change of pH or other conditions \cite{Meirovitch2007}, \cite{Chipot2007}, \cite{Jorgensen2004}. 
Free energy also plays an important role in the case of protein
folding. Ground breaking work of Christian B. Anfinsen and coworkers showed that the native
structures of small globular proteins have a unique, thermodynamically stable native structure with
their conformation at the global free energy minimum \cite{Anfinsen1973}. Regardless of the starting
point or even after unfolding 
it by changing different condition most proteins will go back to the native state after
reinstituting native like conditions. Often, the changes in free energy between different
conformations is small, but nonetheless crucial in favoring the native state
 since defects in protein folding may be
the molecular cause of a range of human genetic disorders. For example a misfolded protein known as
a
prion can misfold correctly folded proteins when entering a healthy organism.
\Alberto{I'm not sure I understand this here} \Arijit{Thus, Free energy can also act as a scale to identify between the misfolded and the native state of
the protein.} These facts further give free energy a special importance in structural biology.

@@

Previous applications of the confiment method have focused on relatively short peptides, up to $17$
residues in length. In this work we apply the confinement method to
larger proteins. On that direction, we first test the confinement method for a chamelon sequence and 
found that it can correctly pick out the prefential structure of the chamelon sequence. Other 
applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimentally determined structure. \Alberto{should not start with results yet, first
we are describing what we are going to show them} As
Table~\ref{table:casp_control} shows, in most cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. 
Most interestingly, we decomposed the free energy into its per residue component, which help 
us to identify the residues which stabilize or destabilize a particular conformation of the protein.
%Although differentiating between experimentally determined structures and computer generated predictions is
%not a stringent test of a scoring method---for example, examining the residue-residue packing can
%easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
%"sanity check".

@@

We decomposed the calculated free energy into per residue decomposition as described in the method
section. This helps us to identify important residues that stabilize a particular conformation. 
In this direction, we first calculate the per residue free energy for the GA95 and GB95 sequences. 
The per residue free energy, $\Delta \Delta G (4 \beta + \alpha$ - $3 \alpha)$ is plotted in Figure~\ref{fig:orban_full}(F).
In this plot, the negative peak indicate the residue will stabilize the $4 \beta + \alpha$ conformation of a particular
residue and the positive peak will stabilize the $3 \alpha$ structure. As, the differences of amino acid
residues are only at 3 position, there are some common features of residues in both GA95 and GB95 sequences in both
$3 \alpha$ and $4 \beta + \alpha$ conformation. Role of some important residues in stabilizing either 
$4 \beta + \alpha$ or $3 \alpha$ conformation and possible reason for such conformational preferences are discussed in 
details in Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}. It is important to note that
experimental observations classified the protein into two parts: Amino acids 9-51 are fully structured in both
the folds, where as residues 1-8 and 52-56 are unstructured in $3 \alpha$, but form $\beta$ strand in $4 \beta + \alpha$ structure.
Most of the amino acid residues in the region 1-9 favor $4 \beta + \alpha$ structure as can be seen from the plot
Figure~\ref{fig:orban_full}(F). Interestingly, we can explain most of the big peak in this plot. They are presented 
in Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}. Here, we explain only one case. The first big peak
appears at position 7. Here, the hydrophobic Leu-7 stabilizes the $4 \beta + \alpha$ conformation as its sidechain is oriented 
towards the protein hydrophobic core, whereas in $3 \alpha$ structure it exposed to the solvent (Figure~\ref{fig:orban_full}(A)).
Thus, if someone want to better design the $3 \alpha$ conformation of this protein, we propose to mutate Leu at this position
with some hydrophilic residue. Similarly, we found that the free energy differences arise in two conformation for various 
reasons ranging from hydrogen/salt bridge formation to relative solvent exposure of hydrophobic residues etc. They are 
listed in Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}.

@@

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test in which different groups world wide apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from better to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}. The consequences of this go beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, from structures for drug design leads to designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. We have tested different options for that purpose.
First, we rank order submition from same group. Then we rank submition from different groups. Our second test was 
to check whether this method can do better quality assessment/metaprediction than the best group in that catagory in CASP9. 
In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 Å).

@@

\subsubsection{Can confinement method perform better quality assessment in protein structure prediction ?}
A part of CASP experiment is dedicated to the assessment of the quality of predicted model \cite{Kryshtafovych2011}. 
Most of the top performing group in this catagory use consensus approaches for quality assessment \cite{Wang2011}.
A review of CASP9 quality assessment also pointed out that the methods which are based on clustering
techniques perform
much better compared to those methos which are based on analysis of individual models \cite{Kryshtafovych2011}. 
In this background we try to answer, whether the confinement method can do better quality assessment compare to the best performing
group (MUFOLD-WQA) in this catagory. In CASP9, the overall reliability of models in quality assessment (QA) was accepted in a mode known as
QMODE 1. Here, predictors were asked to score each model on a scale from 0 to 1, with higher values corresponding to better models \cite{Kryshtafovych2011}. 
We have again picked up the CASP taget T0538 to answer this question. We have chosen the Model 3 submitted by PconsR (GDT\_TS = 96, QMODE 1 = 0.5434) 
and model 5 from the MULTICOM-NOVEL ( GDT\_TS = 83, QMODE 1 = 0.5865). Both are server predicted models and they were further used for quality assessment 
by the group MUFOLD-WQA \cite{Wang2011}. It is important to note that the model from PconsR was the most accurate predicted model for this 
target and the model from MULTICOM-NOVEL was predicted as best by the quality assessment experiments of MUFOLD-WQA. Also, 
the QMODE 1 value presented here is from the group MUFOLD-WQA. 
Our free energy analysis predict that the model from Pconsr is indeed better than the model from 
MULTICOM-NOVEL $(\Delta \Delta G (Model\_Pconsr - Model\_MULTICOM-NOVEL = -3.9 Kcal/Mol)$, which   
matches well with the GDT\_TS score. On the other hand the MUFOLD-WQA predicted the ranking in the reverse order.

@@

all the structures are very very close to the native structure. But that does not matter at all as our main aim is to differentiate
between the good and the bad structure. In this article we have extensively tested this method using examples from CASP experiment.

>>>>

% Custom macros for author comments
\newcommand{\Alberto}[1]{\color{ForestGreen}#1\normalcolor }
\newcommand{\Justin}[1]{\color{blue}#1\normalcolor}
\newcommand{\Arijit}[1]{\color{magenta}#1\normalcolor}
\newcommand{\Ken}[1]{\color{red}#1\normalcolor}

@@

In computational structural biology there are numerous cases where the free energy difference
between two well defined 
states plays an important role. Examples range from small to large conformational changes of proteins due to
ligand binding, change of pH or other conditions \cite{Meirovitch2007}, \cite{Chipot2007}, \cite{Jorgensen2004}. 
Free energy also plays an important role in the case of protein
folding. Ground breaking work of Christian B. Anfinsen and coworkers showed that the native
structures of small globular proteins have a unique, thermodynamically stable native structure with
their conformation at the global free energy minimum \cite{Anfinsen1973}. Regardless of the starting
point or even after unfolding 
it by changing different condition most proteins will go back to the native state after
reinstituting native like conditions. Often, the changes in free energy between different
conformations is small, but nonetheless crucial in favoring the native state
 since defects in protein folding may be the molecular cause of a range of human genetic disorders.
 For example a misfolded protein known as a
prion can misfold correctly folded proteins when entering a healthy organism. Thus it is important for 
protein to have its native structure. As protein native structure stays in the global free energy minimum, 
\Arijit{free energy can be used as a scale for accurate identification of misfolded state from the protein native state.}
\Alberto{I'm not sure I understand this here} 
These facts further give free energy a special importance in structural biology.

@@

Previous applications of the confiment method have focused on relatively short peptides, up to $17$
residues in length. In this work we apply the confinement method to
larger proteins. On that direction, we first test the confinement method for a chamelon sequence and 
found that it can correctly pick out the preferential structure of the chamelon sequence. Other 
applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimentally determined structure. \Alberto{should not start with results yet, first
we are describing what we are going to show them} As
Table~\ref{table:casp_control} shows, in most cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. 
Most interestingly, we decomposed the free energy into its per residue component, which help 
us to identify the residues which stabilize or destabilize a particular conformation of the protein.
%Although differentiating between experimentally determined structures and computer generated predictions is
%not a stringent test of a scoring method---for example, examining the residue-residue packing can
%easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
%"sanity check".

@@

We decomposed the calculated free energy into per residue decomposition as described in the method
section. This helps us to identify important residues that stabilize a particular conformation. 
In this direction, we first calculate the per residue free energy for the GA95 and GB95 sequences. 
The per residue free energy, $\Delta \Delta G ((4 \beta + \alpha$) - ($3 \alpha))$ is plotted in Figure~\ref{fig:orban_full}(F).
In this plot, the negative peak indicate the residue will stabilize the $4 \beta + \alpha$ conformation of a particular
residue and the positive peak will stabilize the $3 \alpha$ structure. As, the differences of amino acid
residues are only at 3 position, there are some common features of residues in both GA95 and GB95 sequences in both
$3 \alpha$ and $4 \beta + \alpha$ conformation. Role of some important residues in stabilizing either 
$4 \beta + \alpha$ or $3 \alpha$ conformation and possible reason for such conformational preferences are discussed in 
details in Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}. It is important to note that
experimental observations classified the protein into two parts: Amino acids 9-51 are fully structured in both
the folds, where as residues 1-8 and 52-56 are unstructured in $3 \alpha$, but form $\beta$ strand in $4 \beta + \alpha$ structure.
Most of the amino acid residues in the region 1-9 favor $4 \beta + \alpha$ structure as can be seen from the plot
Figure~\ref{fig:orban_full}(F). Interestingly, we can explain most of the big peak in this plot. They are presented 
in Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}. Here, we explain only one case. The first big peak
appears at position 7. Here, the hydrophobic Leu-7 stabilizes the $4 \beta + \alpha$ conformation as its sidechain is oriented 
towards the protein hydrophobic core, whereas in $3 \alpha$ structure it exposed to the solvent (Figure~\ref{fig:orban_full}(A)).
Thus, if someone want to better design the $3 \alpha$ conformation of this protein, we propose to mutate Leu at this position
with some hydrophilic residue. Similarly, we found that the free energy differences arise in two conformation for various 
reasons ranging from hydrogen/salt bridge formation to relative solvent exposure of hydrophobic residues etc. They are 
listed in Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}.

@@

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test in which different groups world wide apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from better to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}. The consequences of this go beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, from structures for drug design leads to designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. We have tested different options for that purpose.
First, we rank order submition from same group. Then we rank submition from different groups. Our second test was 
to check whether this method can do better quality assessment/metaprediction than the best group in that catagory in CASP9. 
In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 \AA).

@@

\subsubsection{Can confinement method perform better quality assessment in protein structure prediction ?}
A part of CASP experiment is dedicated to the assessment of the quality of predicted model \cite{Kryshtafovych2011}. 
Most of the top performing group in this catagory use consensus approaches for quality assessment \cite{Wang2011}.
A review of CASP9 quality assessment also pointed out that the methods which are based on clustering
techniques perform
much better compared to those methods which are based on analysis of individual models \cite{Kryshtafovych2011}. 
In this background we try to answer, whether the confinement method can do better quality assessment compare to the best performing
group (MUFOLD-WQA) in this catagory. In CASP9, the overall reliability of models in quality assessment (QA) was accepted in a mode known as
QMODE 1. Here, predictors were asked to score each model on a scale from 0 to 1, with higher values corresponding to better models \cite{Kryshtafovych2011}. 
We have again picked up the CASP taget T0538 to answer this question. We have chosen the Model 3 submitted by PconsR (GDT\_TS = 96, QMODE 1 = 0.5434) 
and model 5 from the MULTICOM-NOVEL ( GDT\_TS = 83, QMODE 1 = 0.5865). Both are server predicted models and they were further used for quality assessment 
by the group MUFOLD-WQA \cite{Wang2011}. It is important to note that the model from PconsR was the most accurate predicted model for this 
target and the model from MULTICOM-NOVEL was predicted as best by the quality assessment experiments of MUFOLD-WQA. Also, 
the QMODE 1 value presented here is from the group MUFOLD-WQA. 
Our free energy analysis predict that the model from Pconsr is indeed better than the model from 
MULTICOM-NOVEL $(\Delta \Delta G (Model\_Pconsr - Model\_MULTICOM-NOVEL = -3.9 Kcal/Mol)$, which   
matches well with the GDT\_TS score. On the other hand the MUFOLD-WQA predicted the ranking in the reverse order.

@@

all the structures are very very close to the native structure. But we want to ignore this fact as our main aim is to differentiate
between the good and the bad structure. In this article we have extensively tested this method using examples from CASP 
